Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-05-03 young_pp_discs.tex
spellcheck etc

{\bf Whilest photochemistry was not included in I2011, it is unclear to what extent the radiation from the protostar will impact upon the disc, given that material infalling from the envelope and newly forming outflows will be obscuring the star. Column densities between upper layers of the disc and the protostar in the model used here are on the order of 10$^{23}$ cm$^{-2}$. This is in contrast to later stages of disc evolution where the envelope has been almost entirely cleared (e.g. Walsh et. al), in this case it is clear that the chemistry in the upper layers of the disc will be dominated by photo-chemical processes.}

@@

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their H$_2$ number density, and the number density of the molecular spieces being simulated, compared against those of a reference point in order to decide if the candidate point is to be used in the grid or not. Candidate grid points are selected at random in a cylindrical coordinates that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

Fig. \ref{continuum} shows the 1\,mm continuum emission and beam size in the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where g is the gas/dust mass ratio S$_\nu$ is the integrated flux, d is the distance to the source, $\kappa_\nu$ is the dust opacity and B$_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. From this, a disc mass measurement of 0.1$\,$M$_\odot$ was obtained by assuming a dust temperature of 40K, and a measurement of 0.005$\,$M$_\odot$ was obtained by using 400K. The flux was integrated from y=0.2 to -0.2 and x=0.4 to -0.4. This shows that the disc is not optically thin to millimetre wavelengths. This underestimate of the disc mass by at least a factor of 4 shows that in young, massive discs such as those we are investigating, longer wavelength continuum observations are neccesary for determining the disc mass.\smallskip

@@

One assumption underpoinning this model is that the gas and dust are in thermal equilibrium in the disc. If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption. However, the large volume densities are expected to provide efficient dust-gas coupling. Outflows could contaminate line profiles of  those species  which commonly trace outflows, such as CO and HCO$^+$. In these cases, a detailed study of their line profiles using high spectral resolution should help to disentangle the various kinematic components.  In the future we will study this, including outflow lobes in the disc/envelope system.  However species such as OCS and C$^{17}$O, which are not commonly seen in outflows, can be used to trace disc rotation and structure.\smallskip

{\bf The disc model used does not include magnetic fields, including magnetic fields may suppress GI and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficicent magnetization and alinegment between the parent cores rotation axis and maxnetic field direction (Joos et al. 2012, Li et al. 2013). This would mean that if observations similar to the ones we have descirbed are made constraints could be placed on B-field strength within cores/discs. \smallskip Future work including looking at the effects of material from the envelope falling onto the disc and how this affects the chemistry and line propagation, and including realistic outflows into the model.}

>>>>

{\bf Whilst photochemistry was not included in I2011, it is unclear to what extent the radiation from the protostar will impact upon the disc, given that material infalling from the envelope and newly forming outflows will be obscuring the star. Column densities between upper layers of the disc and the protostar in the model used here are on the order of 10$^{23}$ cm$^{-2}$ (is this enough to block the protostars visible and UV?). This is in contrast to later stages of disc evolution where the envelope has been almost entirely cleared (e.g. Walsh et. al), in this case it is clear that the chemistry in the upper layers of the disc will be dominated by photo-chemical processes.}

@@

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their H$_2$ number density, and the number density of the molecular species being simulated, compared against those of a reference point in order to decide if the candidate point is to be used in the grid or not. Candidate grid points are selected at random in a cylindrical coordinates that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

Fig. \ref{continuum} shows the 1\,mm continuum emission and beam size in the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where g is the gas/dust mass ratio S$_\nu$ is the integrated flux, d is the distance to the source, $\kappa_\nu$ is the dust opacity and B$_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. From this, a disc mass measurement of 0.1$\,$M$_\odot$ was obtained by assuming a dust temperature of 40K, and a measurement of 0.005$\,$M$_\odot$ was obtained by using 400K. The flux was integrated from y=0.2 to -0.2 and x=0.4 to -0.4. This shows that the disc is not optically thin to millimetre wavelengths. This underestimate of the disc mass by at least a factor of 4 shows that in young, massive discs such as those we are investigating, longer wavelength continuum observations are necessary for determining the disc mass.\smallskip

@@

One assumption underpinning this model is that the gas and dust are in thermal equilibrium in the disc. If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption. However, the large volume densities are expected to provide efficient dust-gas coupling. Outflows could contaminate line profiles of  those species  which commonly trace outflows, such as CO and HCO$^+$. In these cases, a detailed study of their line profiles using high spectral resolution should help to disentangle the various kinematic components.  In the future we will study this, including outflow lobes in the disc/envelope system.  However species such as OCS and C$^{17}$O, which are not commonly seen in outflows, can be used to trace disc rotation and structure.\smallskip

{\bf The disc model used does not include magnetic fields, including magnetic fields may suppress GI and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficient magnetisation and alignment between the parent cores rotation axis and magnetic field direction (Joos et al. 2012, Li et al. 2013). This would mean that if observations similar to the ones we have described are made constraints could be placed on B-field strength within cores/discs. \smallskip Futher work to improve these simulated observations would include looking at the effects of material from the envelope falling onto the disc and how this affects the chemistry and line propagation, and including realistic outflows into the model.}
